

Section 1: Introduction to GGplot2

Examples.

Section 2: The qplot() function

Iris Data Set and Qplot()

Section 7: Faceting


Section 1: Introduction to GGplot2


Introduction to Ggplot2: There are three options available to users.


1-Basic grid graphics installation

2-Lattice graphics

3-Ggplot graphics






http://www.jstatsoft.org/v17/b03/paper

http://www.r-bloggers.com/the-grammar-of-graphics-l-wilkinson/


The syntax is easy to learn, allows the user fine level of control over the output.

Examples.


1 - packages requires

2 - structure histograms scatterplots regression combinations

3 - future projects for Dublin R

4 - Bayesian Stats [ Kevin Burke].

5 - coplot() xyplot(), geom*, .www.ling.upenn.edu 

summer 2010 introduction to GGplot2.


wide variety of graphical displays 


(pg.1)aesthetics positions size of elements shape ofelementscolour of elements.

geometric shapes :Points linea line segments bars text.


(pg.7) by default stat.smooth adds a loess line with standard error represented by a semi transparent ribbon.

The default geoms are geom.ribbon and geom.smooth.


Use points and error bars ( not advised)

(pg.9) stat.summary : takes a few different arguments: fun.y fun.ymax fun.ymin fun.data.


(pg.11) grouping defined as a combination of aesthetics settings or explicitly with the argument grouped.

(pg.19) Positions. How geoms are positioned together may require some adjusting.

Identity, jitter and stack.


(pg. 25) Facet.wrapcreates and labels a plot for every levek of a factor passed to it. The primary arguments take the form of a one sided formula. It will then efficiently wrap these plits into a 2d grid.

Facet scales. Sometimes free scales are appropriate.  

Section 2: The qplot() function









qplot(carat,price,data=dsmall,colour=color)

qplot(carat,price,data=dsmall,shape=cut)

 



This workshop will look at GGplot only (though making comparisons to Base.)

Iris Data Set and Qplot()


consider the famous Iris data set used for statistical teaching. 

qplot()

-----

- geometric objects and statistics

- scales

- themes

- faceting

-----

colour palettes. What selection of colour us available using ggplot2


•
how to create scatterplots 

•
how to add extra variables with . . .



in ggplot you create a plot using ggplot() function.

-----


 - Regression Lines

 - Legends


The fundamental object is known as a Geom.



 


Section 7: Faceting


 Faceting partitions a plot into a matrix of plots.

Use of PANELLLING

TRELLIS plots

subsets of data in different panels


ggplot is a large library includes maps annotations and themes


stat_smooth fits a line [ e.g. Regression fit ] and shaded band to depict uncertainty.


Aes call specifies colour and shape.


How to specify labels per facet in ggplot. 

