
MS4024 Assessment 4b   

 
•
This assessment is worth 10% of the overall mark.  

•
Answer all ten questions. 

•
Each question is worth 1%. 

•
The duration of assessment is 45 minutes.

•
Submit your answers on a PDF document via email to kevin.obrien@ul.ie.

•
Please include all code and plots used in the PDF document when submitting your answers.


 

Resident datasets used : Orange, USArrests, iris

 



--------------------------------------------------------------------------------


 















xdat
 

1.8
 

2.9
 

3.7
 

4.5
 

5.1
 

6.4
 

7.1
 



ydat
 

7.9
 

7.2
 

6.5
 

5.8
 

4.9
 

3.9
 

3.1
 

 

1) What is the mean and standard deviation of both 'xdat' and 'ydat' ?

 

2) What is the correlation coefficient of 'xdat' and 'ydat'?

 

3) From the 'Orange' data set, what is the mean and standard deviation of the 'age' column?

 

4) From the 'USArrests' data, what is the lowest and highest values in the 'Murder' column?

 

5) From the 'iris' data set, what is the mean and standard deviation of petal widths of 'virginica' species subgroup?

 

6) Write a simple function 'Zfunc' to perform the following task:

 

Given three values  (mu),  (sigma) and X, the function will return a value Z , defined as follows.

 

Z=X-

 

The default values are = 0,  = 1 and X = 1 respectively.

 

7) For all species, make a plot of the petal lengths (x-axis) and petal widths (y axis). (plot character: 14 , colour: blue ) 

 

8) Add a sloped line (line type: 2, colour: green) to this plot. The intercept is  1.1 and the slope is 2.23.

 

9) Add the title "Iris data scatterplot" to your plot. 

 

10) Briefly describe what the 'lm' function is used for.

 

 

