Lists
========================================================

- are an ordered collection of components;
-	components may be arbitrary R objects (data frames, vectors,lists, etc. );
- single bracket notation for sublists;
-	double bracket notation for individual components;
-	construct using the function list().

A simple example of a list is as follows:
```{r}
L1 <- list(name="Fred",wife="Mary",no.children=3,
child.ages=c(4,7,9))
```
Each component of the list is given a name (i.e. name, wife, no.children, child.ages).

We can construct a second list omitting the component names:
```{r}
L2 <- list("Fred", "Mary", 3, c(4,7,9))
```
What is the difference between the two? Three equivalent ways of accessing the first component.
```{r}
L1[["name"]]
L1$name
L1[[1]]

```
A sublist consisting of the first component only
> L1[1]
$name
[1] "Fred"

The names of each component of the list can be accessed using
```{r}
names(L1)
names(L2)
```

We can set the names for the list components after the list has been
created.
```{r}
names(L2) <- c("name.hus", "name.wife", "no.child", "child.age")
names(L2)
```

We can also concatenate lists:
```{r}
L3 <- c(L1, L2)
```


